---
title: "NBI_report"
author: "Martin Menninger"
date: "8/16/2019"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(readr)
library(measurements)

setwd("~/R/Federal-Performance-Measures/Bridges")
```

## Download the data

This chunk downloads the data that you need to run the rest of the report. You should only need to run it once. 

```{r download_data, eval=FALSE, include = FALSE}
library(tidyverse)
library(readr)
library(measurements)

##Define years and state you want ############## User Input Here ############
years <- 1992:2018 ##NOTE: Available back to 1992##
state <- "IL"

# Download all the text files for those years
download.file(url = paste("https://www.fhwa.dot.gov/bridge/nbi/", years, "/", state, substring(years, 3), ".txt", sep = ""), 
              destfile = paste("NBI", state, years, ".txt", sep = ""), method = "libcurl")

```


## Import the data

```{r Import into R, include = FALSE, warning=FALSE}
years <- 1992:2018
state <- "IL"

Format <- read_csv("~/R/Federal-Performance-Measures/Bridges/Inputs/Format.csv")
fcol_types <- paste0(rep(Format$Same), collapse = "")

##Needed to set up loop. Should rewrite to eliminate need for this. 
NBIAll<- data.frame(read_fwf(file = paste("NBI", state, "2018", ".txt", sep = ""), 
                           fwf_widths(Format$ITEM_LENGTH, Format$Column_Name), 
                           col_types = fcol_types))
NBIAll$NBI_Year = 2018 #add a column that indicates inventory year

for (i in years){
  NBIAll <- filter(NBIAll, NBIAll$NBI_Year != i) #cleares out old data
  temp <- read_fwf( file = paste("NBI", state, i, ".txt", sep = ""),
                    fwf_widths(Format$ITEM_LENGTH, Format$Column_Name), 
                    col_types = fcol_types)

  temp$NBI_Year = i #add a column that indicates inventory year
  print(i)
  print(Sys.time())
  NBIAll <- bind_rows(NBIAll, temp)
  }

rm(Format)
rm(i, fcol_types, years, state, temp)
#saveRDS(NBIAll, file = "nbiIL92_18.RDS")

```

## Clean Up the Data

```{r clean_data, include = FALSE}
library(readr)
  Owner_code <- read_csv("~/R/Federal-Performance-Measures/Bridges/Inputs/Owner_code.csv")
  condition_levels <- factor( levels = c("Good", "Fair", "Poor"), ordered = TRUE)

  
#Build NBI file ----
  nbi_trend <- NBIAll %>%
    filter(RECORD_TYPE_005A =="1") %>%
    select(STATE_CODE_001, COUNTY_CODE_003, LAT_016, LONG_017, WATERWAY_EVAL_071, HIGHWAY_SYSTEM_104, ## Location
           APPR_WIDTH_MT_032, STRUCTURE_LEN_MT_049, DECK_WIDTH_MT_052,   ##Size
           DECK_COND_058, SUPERSTRUCTURE_COND_059, SUBSTRUCTURE_COND_060, CULVERT_COND_062,  ## Condition
           OWNER_022, FUNCTIONAL_CLASS_026, YEAR_BUILT_027, ADT_029, SUFFICIENCY_RATING, CAT10, NBI_Year) %>%
    left_join(maps::state.fips, by = c("STATE_CODE_001" = "fips")) %>% ## Join in the State
    mutate(county_code = as.integer(paste(STATE_CODE_001, COUNTY_CODE_003, sep = ""))) %>%  ##Format for Join
    left_join(maps::county.fips, by = c("county_code" = "fips")) %>% ## Join in the County 
    mutate(water = ifelse(WATERWAY_EVAL_071 == "N", 0, 1)) %>%
    mutate(age = 2019 - YEAR_BUILT_027) %>%
    left_join(Owner_code, by = c("OWNER_022" = "Code")) %>%
    #mutate(CAT10 = fct_recode(CAT10, "Good" = "G","Fair" = "F","Poor" = "P")) %>%
    mutate(    area_m = ifelse(DECK_WIDTH_MT_052 > 0, #There is an assumed decimal point in len and wdt so /100 
                               STRUCTURE_LEN_MT_049 *DECK_WIDTH_MT_052/100, 
                               STRUCTURE_LEN_MT_049 *APPR_WIDTH_MT_032/100), 
               min_con = pmin(DECK_COND_058,SUPERSTRUCTURE_COND_059, 
                              SUBSTRUCTURE_COND_060, CULVERT_COND_062, na.rm = TRUE), 
               Rating = ifelse(min_con<=4, "Poor", ifelse(min_con>=7, "Good", "Fair")))

#nbi_trend$Rating <- factor(nbi_trend$Rating, levels = condition_levels)  
nbi_trend$Rating <- factor(nbi_trend$Rating, levels = c("Good", "Fair", "Poor"))

#Latitude----  
  deg <- as.numeric(stringr::str_sub(nbi_trend$LAT_016, 1, 2))
  min <- as.numeric(stringr::str_sub(nbi_trend$LAT_016, 3, 4))
  sec <- as.numeric(stringr::str_sub(nbi_trend$LAT_016, 5, 8))/100
  nbi_trend$latitude <- deg + min/60 + sec/3600
  
#Longitude----
  deg <- as.numeric(stringr::str_sub(nbi_trend$LONG_017, 1, 3))
  min <- as.numeric(stringr::str_sub(nbi_trend$LONG_017, 4, 5))
  sec <- as.numeric(stringr::str_sub(nbi_trend$LONG_017, 6, 9))/100
  
  nbi_trend$longitude <- -1 * (deg + min/60 + sec/3600)
 
  nbi <- nbi_trend %>% filter(NBI_Year == max(nbi_trend$NBI_Year))
  
   nbi <- nbi %>%
    filter(longitude > -92) %>%
    filter(latitude > 36) %>%
    filter(latitude < 50)

rm(condition_levels, deg, min, sec) 
rm(Owner_code)
#rm(NBIAll)
rm(temp)

```

## Add geography information for summary

This links the Illinois bridge data to the CMAP council of Mayors. If you do not live in the CMAP region, you probably want to skip this step. 

```{r Download the Council of Mayor Geography}
library(sf)
library(sp)
#Download Council of Mayoors  (only once)
COM_path <- "https://datahub.cmap.illinois.gov/dataset/a466a0bf-3c3e-48cb-8297-6eb55f49fb50/resource/cbe742e0-2e78-4a54-95c2-d694703e1ae6/download/CoMCMAP201303.zip"
download.file(COM_path, "COM_2013.zip")
unzip("COM_2013.zip", exdir = ".")
COM <- st_read("CoM_CMAP_201303.shp", stringsAsFactors = FALSE)    #Local Council of Mayors
plot(COM)
```



```{r Join COM to bridges}
#Make the NBI spatial
nbisp <- subset(nbi_trend, is.na(latitude) == F)
nbisp <- subset(nbisp, is.na(longitude) == F)
#nbisp<- filter(nbisp, COUNTY_CODE_003 == "031" )
#coordinates(nbisp) <- c("longitude", "latitude")
nbisp<- st_as_sf(nbisp, coords = c("longitude", "latitude"))
nbisp <-st_set_crs(nbisp,  "+proj=longlat +ellps=GRS80 +datum=NAD83 +no_defs")
#nbisp<- st_as_sf(nbisp)
nbisp<- st_transform(nbisp, st_crs(COM))
nbisp <- st_join(nbisp, COM)


#plot(subset(z, z$Council=="City of Chicago"))

```

## Bridge Condition Trends

The federal performance measure focuses on bridges on the NHS. Note that the federal goal is to have less than 10%.  

```{r State_NHS trend}
nbi_trend %>% 
  filter(HIGHWAY_SYSTEM_104 == 1) %>% #Filter for NHS only
  filter(Rating != "NA") %>% #Remove bad data (need to inspec errors)
  group_by(NBI_Year, Rating) %>%
  summarise(countt = n(), area = sum(area_m)) %>%
  mutate(pct = area/sum(area)) %>%
  ggplot(aes(x= NBI_Year, y = pct*100, fill = Rating)) + 
  geom_bar(stat = "identity") +
  #geom_text(aes(label = paste(format(pct*100, digits = 1), "%")),
  geom_text(aes(label = paste(round(pct*100, digits = 1), "%", sep = "")), 
            position = position_stack(vjust = 0.5), size = 3, angle = 90) +
  scale_x_continuous(breaks = seq(1992, 2018, 2), limits = c(1990, 2020))+
  scale_fill_brewer(type = "seq", palette="Spectral", direction = -1) +
  #coord_flip() +
  labs(x= "Year", y= "Percent of Bridges", title= "Percent of IL NHS Bridges Area By Condition") 
```


## CMAP performance

Note that CMAP increased past the federal goal of 10% in 2018. CMAP has a local goal for 2021 of At least 36.8% good and no more than 8.3% poor. 

```{r CMAP_NHS_Trend}
  nbisp %>%
  filter(HIGHWAY_SYSTEM_104 == 1) %>%
  #filter(COUNTY_CODE_003 %in% c("031", "043", "089", "093", "097", "111", "197")) %>% #Limit to CMAP counties
  #filter(Rating != "NA") %>%
  filter(Council != "NA") %>%   #A filter for CMAP area - broken for years 02 and 13 with bad lat lon. 
  group_by(NBI_Year, Rating) %>%
  summarise(countt = n(), area = sum(area_m)) %>%
  #group_by(NBI_Year, Council) %>%
  mutate(pct = area/sum(area)) %>%
  
  
  ggplot( aes(x= NBI_Year, y = pct*100, fill = Rating)) + 
  geom_bar(stat = "identity") +
  #geom_text(aes(label = paste(format(pct*100, digits = 1), "%")),
  geom_text(aes(label = paste(round(pct*100, digits = 1), "%", sep = "")), 
            position = position_stack(vjust = 0.5), size = 3, angle = 90) +
  scale_x_continuous(breaks = seq(1992, 2018, 2), limits = c(1990, 2020))+
    scale_fill_brewer(type = "seq", palette="Spectral", direction = -1) +
  #facet_grid(rows = vars(Council)) +
  #facet_wrap(~Council) +
  #coord_flip() +
  labs(x= "Year", y= "Percent of Bridges", title= "Percent of CMAP NHS Bridges Area By Condition")

```

## Other measures

While CMAP tracks the federal performance target, we also care about bridges not on the NHS. Note that non-NHS bridges are generally in worse condition. 

```{r State_non-NHS_trend}
nbi_trend %>% 
  filter(HIGHWAY_SYSTEM_104 == 0) %>%
  filter(Rating != "NA") %>%
  filter(COUNTY_CODE_003 %in% c("031", "043", "089", "093", "097", "111", "197")) %>% #Limit to CMAP counties
  group_by(NBI_Year, Rating) %>%
  summarise(countt = n(), area = sum(area_m)) %>%
  mutate(pct = area/sum(area)) %>%
  ggplot(aes(x= NBI_Year, y = pct*100, fill = Rating)) + 
  geom_bar(stat = "identity") +
  #geom_text(aes(label = paste(format(pct*100, digits = 1), "%")),
  geom_text(aes(label = paste(round(pct*100, digits = 1), "%", sep = "")), 
            position = position_stack(vjust = 0.5), size = 3, angle = 90) +
  scale_x_continuous(breaks = seq(1992, 2018, 2), limits = c(1990, 2020))+
  scale_fill_brewer(type = "seq", palette="Spectral", direction = -1) +
  #coord_flip() +
  labs(x= "Year", y= "Percent of Bridges", title= "Percent of CMAP non-NHS Bridges Area By Condition") 
```


## Councils

Some federal funding is sub allocated by local Council of Mayors. This chart shows the condition over time of the bridges that are maintained with those funds. 

```{r Council_Chart, fig.width = 8, fig.height=20}
  nbisp %>%
  #filter(HIGHWAY_SYSTEM_104 == 1) %>% # We dont care what system they are on
  #filter(COUNTY_CODE_003 %in% c("031", "043", "089", "093", "097", "111", "197")) %>%
  filter(OWNER_022 %in% c('02','03','04')) %>% # These are the bridges funded with STP-L
  filter(Rating != "NA") %>%
  group_by(NBI_Year, Rating, Council) %>%
  summarise(countt = n(), area = sum(area_m)) %>%
  group_by(NBI_Year, Council) %>%
  mutate(pct = area/sum(area)) %>%
  filter(Council != "NA") %>%
  
  ggplot( aes(x= NBI_Year, y = pct*100, fill = Rating)) + 
  geom_bar(stat = "identity") +
  #geom_text(aes(label = paste(format(pct*100, digits = 1), "%")),
  geom_text(aes(label = paste(round(pct*100, digits = 1), "%", sep = "")), 
            position = position_stack(vjust = 0.5), size = 2.5) +
  scale_x_continuous(breaks = seq(1992, 2018, 2), limits = c(1990, 2020))+
    scale_fill_brewer(type = "seq", palette="Spectral", direction = -1) +
  facet_grid(rows = vars(Council)) +
  #facet_wrap(~Council) +
  #coord_flip() +
  labs(x= "Year", y= "Percent of Bridges", title= "Percent of Bridges Area By Condition and Council of Mayors") 

```

Funding is allocated based on the square feet of bridge area in Poor condition. 

```{r Council_Chart, fig.width = 8, fig.height=20}
  nbisp %>%
  #filter(HIGHWAY_SYSTEM_104 == 1) %>% # We dont care what system they are on
  #filter(COUNTY_CODE_003 %in% c("031", "043", "089", "093", "097", "111", "197")) %>%
  filter(OWNER_022 %in% c('02','03','04')) %>% # These are the bridges funded with STP-L
  filter(Rating != "NA") %>%
  group_by(NBI_Year, Rating, Council) %>%
  summarise(countt = n(), area = sum(area_m)) %>%
  group_by(NBI_Year, Council) %>%
  mutate(pct = area/sum(area)) %>%
  filter(Council != "NA") %>%
  filter(Rating == "Poor")%>%  #Remove to get all condition
  
  ggplot( aes(x= NBI_Year, y = area*10.7639, fill = Rating)) + 
  geom_bar(stat = "identity") +
  #geom_text(aes(label = paste(format(pct*100, digits = 1), "%")),
  geom_text(aes(label = paste(round(area*10.7639, digits = 0), " sq ft", sep = ""), angle = 90), 
            position = position_stack(vjust = 0.5), size = 2.5) +
  scale_x_continuous(breaks = seq(1992, 2018, 2), limits = c(1990, 2020))+
    scale_fill_brewer(type = "seq", palette="Spectral", direction = -1) +
  facet_grid(rows = vars(Council), scales = "free_y") +
  #facet_wrap(~Council) +
  #coord_flip() +
  labs(x= "Year", y= "Percent of Bridges", title= "Square Ft. of Poor Bridge Area by Council of Mayors",                       subtitle = "Note different scales, 2002 and 2013 data removed") 

```

Sample maps showing most bridges that are not over water, are where we know there are expressways. 

```{r ILBridgeMap}

 nbi %>% 
  filter(longitude < -75) %>%
  filter(COUNTY_CODE_003 %in% c("031", "043", "089", "093", "097", "111", "197")) %>% #CMAP Counties
  mutate(water = as.character(water)) %>%

ggplot()+
  #geom_sf(data = COM) +
  geom_point(aes( x= longitude, y = latitude, color = water), size = .1)+
  scale_color_manual(values = c("red", "blue")) +
  coord_quickmap()
  
```
